---
title: "Hands-On Exercise 10: Information Dashboard Design"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

The following R packages will be required.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse)
```

# 2. Importing Data

We will be using a personal database in Microsoft Access mdb format called *Coffee Chain*.

The `odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package is used to import a database query table into R.

```{r}
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

# 3. Data Preparation

The code chunk below aggregates Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
    group_by(`Product`) %>%
    summarise(`target` = sum(`Budget Sales`),
              `current` = sum(`Sales`)) %>%
    ungroup()
```

# 4. Bullet chart in ggplot2

```{r}
#| code-fold: true
#| code-summary: "Bullet Chart with ggplot2 functions"
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

# 5. Plotting Sparklines using ggplot2

## 5.1 Preparing Data

```{r}
#| code-fold: true
#| code-summary: "Deriving the sum of sales by product and month"
sales_report <- coffeechain %>%
    filter(Date >= "2013-01-01") %>%
    mutate(Month = month(Date)) %>%
    group_by(Month, Product) %>%
    summarise(Sales = sum(Sales)) %>%
    ungroup() %>%
    select(Month, Product, Sales)
```

Next, we compute the minimum, maximum and end of the month sales.

```{r}
#| code-fold: true
#| code-summary: "Compute minimum, maximum, end of month sales"
mins <- group_by(sales_report, Product) %>% 
    slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
    slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
    filter(Month == max(Month))
```

```{r}
#| code-fold: true
#| code-summary: "Compute 25 and 75 quantiles"
quarts <- sales_report %>%
    group_by(Product) %>%
    summarise(quart1 = quantile(Sales, 0.25),
              quart2 = quantile(Sales, 0.75)) %>%
    right_join(sales_report)
```

## 5.2 Sparklines in ggplot2

```{r}
#| code-fold: true
#| code-summary: "Plotting Sparklines"
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

# 6. Static Information Dashboard Design: *gt* and *gtExtras* Methods

-   [gt](https://gt.rstudio.com/) package

-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) package

## 6.1 Plotting a Simple Bullet Chart

```{r}
#| code-fold: true
#| code-summary: "Prepare a Bullet Chart Report"
library(svglite)
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# 7. Sparklines: gtExtras Method

```{r}
#| code-fold: true
#| code-summary: "Data Preparation"
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

We need to convert the report data.frame into list columns as part of the `gtExtras` functions requirements.

```{r}
#| code-fold: true
#| code-summary: "Converting to list columns"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## 7.1 Plotting Coffeechain Sales Report

```{r}
#| code-fold: true
#| code-summary: "Plotting sparkline"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

## 7.2 Adding Statistics

```{r}
#| code-fold: true
#| code-summary: "Calculate summary statistics"
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

## 7.3 Combining the data.frame

```{r}
#| code-fold: true
#| code-summary: "Add statics on the tables"
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
#| code-fold: true
#| code-summary: "Joining the 2 tables"
sales_data = left_join(sales, spark)
```

## 7.4 Plotting the Updated data.table

```{r}
#| code-fold: true
#| code-summary: "Plotting sales_data sparklines"
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## 7.5 Combining Bullet Chart and Sparklines

```{r}
#| code-fold: true
#| code-summary: "Prepare the relevant tables"
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 

sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
#| code-fold: true
#| code-summary: "Combine bullet chart and sparklines"
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# 8. Interactive Information Dashboard Design: *reactable* and *reactablefmtr* methods

-   [reactable](https://glin.github.io/reactable/index.html)

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html)

```{r}
#| code-fold: true
#| code-summary: "Install and load dataui R package"
remotes::install_github("timelyportfolio/dataui")
library(dataui)
```

## 8.1 Plotting Interactive Sparklines

```{r}
#| code-fold: true
#| code-summary: "Plotting with react_sparkline()"
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))

reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## 8.2 Changing Pagesize

```{r}
#| code-fold: true
#| code-summary: "defaultPageSize to change the default setting of pagesize 10"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## 8.3 Adding Points and Labels

-   `highlight_points` argument shows the minimum and maximum value points

-   `label` argument labels the first and last values

```{r}
#| code-fold: true
#| code-summary: "Showing first, last, min and max values"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## 8.4 Adding Reference Line

-   `statline` argument shows the mean line

```{r}
#| code-fold: true
#| code-summary: "Showing the mean line"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## 8.5 Adding Bandline

-   `bandline` argument to add bandline

```{r}
#| code-fold: true
#| code-summary: "Showing the bandline"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## 8.6 Changing from sparkline to sparkbar

```{r}
#| code-fold: true
#| code-summary: "Showing the sparkbars"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

[^1]

[^1]: This document was completed with reference to:

    -   Prof Kam's notes "Information Dashboard Design: R methods": <https://r4va.netlify.app/chap31>
