---
title: "Hands-On Exercise 08c: Analytical Mapping"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

Let's install the required R packages.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# 2. Importing Data

We will load the *NGA_wp.rds* dataset with `read_rds()`.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

# 3. Basic Choropleth Mapping

Let's visualise the distribution of non-functional waterpoint.

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)
```

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total  water point by LGAs",
            legend.outside = FALSE)
```

Displaying the 2 graphs side-by-side:

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

# 4. Choropleth Map for Rates

Lets tabulate the proportion of functional water points and the proportion of non-functional water points in each LGA.

The code chunk below derives `pct_functional` and `pct_nonfunctional` with `mutate()` from dplyr package.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

Next, we plot the map of rate.

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

# 5. Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper end of the scale, with the goal of identifying outliers. 

## 5.1 Percentile Map

The percentile map is a special type of quantile map with six specific categories: 0-1%, 1-10%, 10-50%, 50-90%, 90-99% and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1). Note that the begin and endpoint need to be included.

### Step 1: Exclude records with NA

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

### Step 2: Create customised classification and extract values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

### Step 3: Create *get.var* function

First, we write an R function to extract a variable (i.e. *wp_nonfunctional*) as a vector out of an sf data.frame.

-   arguments:

    -   `vname`: variable name (as character, in quotes)

    -   `df`: name of sf data frame

-   returns:

    -   `v`: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### Step 4: Percentile mapping function

Next, we write a percentile mapping function.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

### Step 5: Test the percentile mapping function

```{r}
percentmap("total_wp", NGA_wp)
```

## 5.2 Box Map

We display the summary statistics on a choropleth map using the basic principles of boxplot.

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

We create break points for a box map.

-   arguments:

    -   `v`: vector with observations

    -   `mult`: multiplier for IQR (default 1.5)

-   returns:

    -   `bb`: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

We create `get.var` function to extract a variable as a vector out of an sf data frame.

-   arguments:

    -   `vname`: variable name (as character, in quotes)

    -   `df`: name of sf data frame

-   returns:

    -   `v`: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

```{r}
# Testing the newly created function

var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

Lastly, lets create a box map.

-   arguments:

    -   `vnam`: variable name (as character, in quotes)

    -   `df`: simple features polygon layer

    -   `legtitle`: legend title

    -   `mtitle`: map title

    -   `mult`: multiplier for IOR

-   returns

    -   a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
