---
title: "Hands-On Exercise 08a: Choropleth Mapping with R"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

The following R packages will be used:

-   [**readr**](https://readr.tidyverse.org/): importing delimited text file

-   [**tidyr**](https://tidyr.tidyverse.org/): tidying data

-   [**dplyr**](https://dplyr.tidyverse.org/): wrangling data

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html): handling geospatial data

    -   [Simple Features for R](https://r-spatial.github.io/sf/)

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# 2. Importing Data into R

## 2.1 Importing Geospatial Data

The code chunk below uses the `st_read()` function of *sf* package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")

# layer name is without the extension

mpsz
```

## **2.2 Importing Attribute Data into R**

The code chunk below imports the *respopagsex2011to2020.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

# 3. Data Preparation

We need to prepare a data table with year 2020 values before creating a thematic map. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY:

-   YOUNG: age group 0 to 4 until age groyup 20 to 24

    ECONOMY ACTIVE: age group 25-29 until age group 60-64

-   AGED: age group 65 and above

-   TOTAL: all age group

-   DEPENDENCY: the ratio between young and aged against economy active group

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

We need to convert the values in PA and SZ fields to uppercase, as the values in these fields are made up of upper and lowercase. On the other hand, the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

`left_join()` of *dplyr* is used to join the geographical data and attribute table using planning subzone name e.g. `SUBZONE_N` and `SZ` as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
`left_join()` of *dplyr* package is used with `mpsz` simple feature data frame as the left data table is to ensure that the output will be a simple features data frame.
:::

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")

# save the dataset that has already been tidied
# can retrieve rds data faster
```

# 4. Choropleth Mapping Geospatial Data Using *tmap*

## 4.1 Plotting quickly using *qtm()*

The easiest and quickest to draw a choropleth map using *tmap* is using `qtm()`. It is concise and provides a good default visualisation in many cases.

The code chunk below will draw a cartographic standard choropleth map as shown below.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

::: callout-note
-   `tmap_mode()` with “plot” option is used to produce a static map. For interactive mode, “view” option should be used.

-   `fill` argument is used to map the attribute (i.e. DEPENDENCY)
:::

## 4.2 Plotting using *tmap*'s elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                  style = "quantile",
                  n = 5,
                  values = "brewer.blues"),
              fill.legend = tm_legend(
                  title = "Dependency ratio")) +
    tm_title("Distribution of Dependency Ratio by planning subzone") +
    tm_layout(frame = TRUE) + 
    tm_borders(fill_alpha = 0.5) + 
    tm_compass(type = "8star", size = 2) +
    tm_grid(alpha = 0.2) +
    tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
               position = c("left", "bottom"))
```

### 4.2.1 Drawing a base map

In the code chunk below, `tm_shape()` is used to define the input data (i.e *mpsz_pop2020*) and `tm_polygons()` is used to draw the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### 4.2.2 Drawing a choropleth map using *tm_polygons()*

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we just need to assign the target variable such as `Dependency` to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

::: callout-note
-   The default interval binning used to draw the choropleth map is called “pretty”. A detailed discussion of the data classification methods supported by **tmap** will be provided in sub-section 4.3.

-   The default colour scheme used is `YlOrRd` of ColorBrewer. You will learn more about the color scheme in sub-section 4.4.

-   By default, missing values will be shaded in grey.
:::

### 4.2.3 Drawing a choropleth map using *tm_fill()* and *tm_border()*

`tm_polygons()` is a wraper of `tm_fill()` and `tm_border()`. `tm_fill()` shades the polygons by using the default colour scheme and `tm_borders()` adds the borders of the shapefile onto the choropleth map.

The code chunk below draws a choropleth map by using `tm_fill()` alone.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

`tm_borders()` is used in the code chunk below to add the boundary of the planning subzones.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

::: callout-note
-   Light-gray border lines have been added on the choropleth map.

-   The `alpha` argument is used to define transparency number between 0 (totally transparent) and 1 (not transparent).

    -   By default, the alpha value of the col is used (normally 1).

-   Beside `alpha` argument, there are three other arguments for `tm_borders()`, they are:

    -   `col` = border colour,

    -   `lwd` = border line width. The default is 1, and

    -   `lty` = border line type. The default is “solid”
:::

## 4.3 Data classification methods of tmap

*tmap* provides a total ten data classification methods, namely: *fixed* *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, and *jenks*.

To define a data classification method, the `style` argument of `tm_fill()` or `tm_polygons()` will be used.

### 4.3.1 Plotting choropleth maps with built-in classification methods

The code chunk below shows a quantile data classification that used 5 classes.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

The code chunk below uses `equal` data classification method.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

::: panel-tabset
#### n = 2

```{r}
#| echo: FALSE
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### n = 6

```{r}
#| echo: FALSE
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### n = 10

```{r}
#| echo: FALSE
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### n = 20

```{r}
#| echo: FALSE
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "equal") +
  tm_borders(alpha = 0.5)
```
:::

### 4.3.2 Plotting choropleth map with custom break

Let's get some descriptive statistics on the variable before setting the break points. The code chunk below computes and displays the descriptive statistics of `DEPENDENCY` field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

we set break point at 0.60, 0.70, 0.80 and 0.90. In addition, we also need to include a minimum and maximum, which we set at 0 and 100. Our `breaks` vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

## 4.4 Colour Scheme

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

Reverse the colour shading by adding a "-" prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

## 4.5 Map Layouts

### 4.5.1 Map Legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

### 4.5.2 Map style

*tmap* allows a wide variety of layout settings to be changed. They can be called by using `tmap_style()`.

The code chunk below shows the `classic` style is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### 4.5.3 Cartographic Furniture

In the code chunk below, `tm_compass()`, `tm_scale_bar()` and `tm_grid()` are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

\
The code chunk below resets the default style.

```{r}
tmap_style("white")
```

## 4.6 Drawing Small Multiple Choropleth Maps

In *tmap*, small multiple maps can be plotted in three ways:

-   by assigning multiple values to at least one of the asthetic arguments

-   by defining a group-by variable in `tm_facets()`

-   by creating multiple stand-alone maps with `tmap_arrange()`

### 4.6.1 Assign multiple values to at least one of the aesthetic arguments

The code chunk below created small multiple choropleth maps with `ncols` in `tm_fill()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

The code chunk below creates small multiple choropleth maps by assigning multiple values to at least one of the aesthetic arguments.

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### 4.6.2 Defining a group-by variable in *tm_facets()*

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### 4.6.3 Create multiple stand-alone maps with *tmap_arrange()*

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## 4.7 Mappping Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
