---
title: "Hands-On Exercise 07: Visualising and Analysing Time-oriented Data"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

The following R packages will be used:

-   scales

-   viridis

-   lubridate

-   ggthemes

-   gridExtra

-   readxl

-   knitr

-   data.table

-   CGPfunctions

-   ggHoriPlot

-   tidyverse

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes, gridExtra, 
               readxl, knitr, data.table, CGPfunctions, 
               ggHoriPlot, tidyverse)
```

# 2. Plotting Calendar Heatmap

We will plot a calender heatmap programmatically using ggplot2 package.

## 2.1 Importing data

The code chunk below imports the *eventlog.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
attacks <- read_csv("data/eventlog.csv")
```

Examine the dataframe before further analysis with `summary()` and `kable()`. `kable()` allows us to review the data structure.

```{r}
summary(attacks)
```

```{r}
kable(head(attacks))
```

There are 3 columns:

-   *timestamp:* stores date-time values in POSIXct format.

-   *source_country:* stores the source of the attack. It is in *ISO 3166-1 alpha-2* country code.

-   *tz:* stores time zone of the source IP address.

## 2.2 Data preparation

### 2.2.1 Derive *weekday* and *hour of day*

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

::: callout-note
-   [`ymd_hms()`](https://lubridate.tidyverse.org/reference/ymd_hms.html) and [`hour()`](https://lubridate.tidyverse.org/reference/hour.html) are from the [lubridate](https://lubridate.tidyverse.org/) package

-   [`weekdays()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/weekdays) is a base R function
:::

### 2.2.2 Derive *attacks* tibble data frame

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

::: callout-note
`mutate()` of **dplyr** package is used to convert *wkday* and *hour* fields into **factor** so they’ll be ordered when plotting.
:::

## 2.3 Building the calendar heatmaps

```{r}

grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "Helvetica") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) ) 
```

::: callout-note
-   `na.omit()`: exclude missing values

-   `geom_tile()`: plot tiles (grids) at each x and y position

-   `color` and `size`: specify the border color and line size of the tiles

-   [`theme_tufte()`](https://jrnold.github.io/ggthemes/reference/theme_tufte.html) of [**ggthemes**](https://jrnold.github.io/ggthemes/reference/index.html) package: remove unnecessary chart junk

-   `coord_equal()`: ensure the plot will have an aspect ratio of 1:1

-   `scale_fill_gradient()`: create a two colour gradient (low-high)
:::

## 2.4 Building Multiple Calendar Heatmaps

First, we identify the top 4 countries with the highest number of attacks, using `count()` and `top_n()`.

Next, we filter the dataset with only the identified countries using `filter()`.

Then, we group the data with an additional variable, `source_country`.

```{r}
top_countries <- attacks %>%
  count(source_country, sort = TRUE) %>%
  top_n(4, n) %>%
  pull(source_country)

filtered_attacks <- attacks %>%
  filter(source_country %in% top_countries)

grouped <- filtered_attacks %>% 
  count(source_country, wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "Helvetica") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) ) +
facet_wrap(~ source_country)
```

## 2.5 Solution - Plotting Multiple Calendar Heatmaps

### 2.5.1 Deriving attack by country object

Identify the top 4 countries with the highest number of attacks:

1.  count the number of attacks by country

2.  calculate the percent of attackes by country

3.  save the results in a tibble data frame

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

### 2.5.2 Preparing the tidy dataframe

Extract the attack records of the top 4 countries from the *attacks* dataframe.

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

### 2.5.3 Plotting using ggplot2 package

```{r}
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          size = 0.1) + 
  theme_tufte(base_family = "Helvetica") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )
```

# 3. Plotting Cycle Plot

## 3.1 Importing data

The code chunk below imports the *arrivals_by_air.xlsx* file into R using the `read_excel()` function of *readxl* package.

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

## 3.2 Deriving *month* and *year* fields

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

## 3.3 Extract the target country

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```

## 3.4 Computing year average arrivals by month

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

## 3.5 Plotting the cycle plot

```{r}
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             size=0.5) + 
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

# 4. Plotting Slopegraph

We will plot a [slopegraph](https://www.storytellingwithdata.com/blog/2020/7/27/what-is-a-slopegraph) using R.

## 4.1 Importing data

The code chunk below imports the *rice.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
rice <- read_csv("data/rice.csv")
```

## 4.2 Plotting the slopegraph

```{r}
rice %>% 
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption = "Prepared by: Dr. Kam Tin Seong")
```

::: callout-note
For effective data visualisation design, `factor()` is used convert the value type of *Year* field from numeric to factor.
:::

[^1]

[^1]: This document was completed with reference to:

    -   Prof Kam's notes "17 Visualising and Analysing Time-oriented Data": <https://r4va.netlify.app/chap17>
