---
title: "Hands-on Ex05d: Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

We will load `GGally`, `parcoords`, `parallelPlot` and `tidyverse` to R.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

The code chunk below imports the *WHData-2018.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

# 2. Plotting Static Parallel Coordinates Plot

`ggparcoord()` of GGally package is used to plot static parallel coordinates plot.

## 2.1 Plotting a simple parallel coordinates

The code chunk below plots a basic static parallel coordinates plot using `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

::: callout-note
Only `data` and `columns` argument are used. `data` is used to map the data object (i.e. *wh*) and `columns` is used to select the columns for preparing the parallel coordinates plot.
:::

## 2.2 Plotting a parallel coordinates with boxplot

The basic parallel coordinates is not meaningful in helping to understand the World Happiness measures.

The code chunk below provides further insights by adding a boxplot in the orginal plot.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

::: callout-note
The following arguments are used:

-   `groupColumn`: group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name

-   `scale`: scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one

-   `alphaLines`: reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1

-   `boxplot`: turn on the boxplot with `TRUE`. The default value is `FALSE`

-   `title`: provide the parallel coordinates plot a title
:::

## 2.3 Parallel coordinates with facet

In the code chunk below, `facet_wrap()` of *ggplot2* is used to plot 10 small miltiple parallel coordinates plots.

Each plot represents 1 geographical region.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

## 2.4 Rotating x-axis text label

The x-axis label can be rotated to make it easier to read, using `theme()` function.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

::: callout-note
-   `axis.text.x` is used to rotate the x-axis text labels

-   `element_text(angle=30)` is used to rotate the text by an angle of 30 degree
:::

## 2.5 Adjusting the rotated x-axis text label

`hjust` argument can be used within `element_text()` to adjust the text location so the label does not overlap with the plot.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# 3. Plotting Interactive Parallel Coordinates Plot: *parallelPlot* methods

The *parallelPlot* package allows us to build interactive parallel coordinates plot.

## 3.1 The basic plot

The code chunk below plots an interactive parallel coordinates plot using `parallelPlot()`.

```{r}
wh_p <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

Note that some of the axis labels are too long.

## 3.2 Rotate the axis label

`rotateTitle` argument can be used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

## 3.3 Changing the colour scheme

`continuousCS` argument can be used to change the default blue colour scheme.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## 3.4 Parallel coordinates plot with histogram

`histoVisibility` argument can be used to plot histogram along the axis of each variable. This provides the distribution of each variable.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

[^1]

[^1]: This document was completed with reference to:

    -   Prof Kam's notes "15 Visual Multivariate Analysis with Parallel Coordinates Plot": <https://r4va.netlify.app/chap15>
