---
title: "Hands-On Exercise 05a: Creating Ternary Plot with R"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

2 main R packages will be used:

-   [**ggtern**](http://www.ggtern.com/): a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/): an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js. The *plotly R* libary contains the *ggplotly* function, which will convert *ggplot2* figures into a Plotly object.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

::: callout-note
V3.2.1 of *ggplot2* will be installed instead of the latest version of *ggplot2*. This is because the current version of *ggtern* package is not compatible to the latest version of *ggplot2*.
:::

The code chunk below imports the *respopagsex2000to2018_tidy.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

We use the `mutate()` function of *dplyr* package to derive three new measures: young, active, and old.

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)

agpop_mutated
```

# 2. Plotting Ternary Diagram with R

## 2.1 Plotting a static ternary diagram

Use `ggtern()` function of *ggtern* package to create a simple static ternary plot.

::: panel-tabset
#### The plot

```{r}
#| echo: FALSE
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

#### The code

```{r}
#| eval: FALSE
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```
:::

The code chunk below adds label and a theme with colours to the plot.

::: panel-tabset
#### The plot

```{r}
#| echo: FALSE
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

#### The code

```{r}
#| eval: FALSE
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```
:::

## 2.2 Plotting an interactive ternary diagram

The code chunk below creates an interactive ternary plot using `plot_ly()` function of *Plotly R*.

::: panel-tabset
#### The plot

```{r}
#| echo: FALSE
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

#### The code

```{r}
#| eval: FALSE
# Reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# Reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
:::

[^1]

[^1]: This document was completed with reference to:

    -   Prof Kam's notes "13 Creating Ternary Plot with R": <https://r4va.netlify.app/chap13>
