---
title: "Hands-On Exercise 09: Modelling, Visualising and Analysing Network Data with R"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

The following R packages will be required.

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts)
```

# 2. Importing Data

There are 2 datasets: the edges (links) data and nodes data.

## 2.1 The Edges Data

The code chunk below imports the *GAStech-email_edges.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## 2.2 The Nodes Data

The code chunk below imports the *GAStech-email_nodes.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
```

## 2.3 Reviewing Data

```{r}
glimpse(GAStech_edges)
```

```{r}
glimpse(GAStech_nodes)
```

## 2.4 Wrangling Time

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

```{r}
glimpse(GAStech_edges)
```

::: callout-note
-   `dmy()` and `wday()` are functions of [lubridate](https://r4va.netlify.app/cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) package.

-   `dmy()` transforms the SentDate to Date data type.

-   `wday()` returns the day of the week as a decimal number or an ordered factor if label is TRUE. The argument abbr is FALSE keep the data spells in full, e.g. Monday.

-   The values in the *Weekday* field are in ordinal scale.
:::

## 2.5 Wrangling Attributes

We aggregate the individual by date, senders, receivers, main subject and day of week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

```{r}
glimpse(GAStech_edges_aggregated)
```

::: callout-note
-   Four functions from **dplyr** package are used: `filter()`, `group()`, `summarise()`, and `ungroup()`.

-   A new field called `Weight` has been added in GAStech_edges_aggregated.
:::

# 3. Create Network Object with Tidygraph

The [tidygraph](https://www.data-imaginist.com/2017/introducing-tidygraph/) package provides a tidy API for graph/network manipulation. 2 functions of the package can be used to create network objects:

-   [`tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) creates a **tbl_graph** network object from nodes and edges data.

-   [`as_tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) converts network data and objects to a **tbl_graph** network. Below are network data and objects supported by `as_tbl_graph()`:

    -   a node data.frame and an edge data.frame,

    -   data.frame, list, matrix from base,

    -   igraph from igraph,

    -   network from network,

    -   dendrogram and hclust from stats,

    -   Node from data.tree,

    -   phylo and evonet from ape, and

    -   graphNEL, graphAM, graphBAM from graph (in Bioconductor)

## 3.1 Using *`tbl_graph()`* to Build Tidygraph Data Model

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)

GAStech_graph
```

::: callout-note
-   The output reveals that *GAStech_graph* is a tbl_graph object with 54 nodes and 4541 edges.

-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.

-   It states that the Node Data is **active**. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time.
:::

## 3.2 Changing the Active Object

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

# 4. Plotting Static Network Graphs with *ggraph* Package

[**ggraph**](https://ggraph.data-imaginist.com/) is an extension of **ggplot2**. There are three main aspects to a **ggraph**’s network graph:

-   [nodes](https://cran.r-project.org/web/packages/ggraph/vignettes/Nodes.html),

-   [edges](https://cran.r-project.org/web/packages/ggraph/vignettes/Edges.html)

-   [layouts](https://cran.r-project.org/web/packages/ggraph/vignettes/Layouts.html)

## **4.1 Plotting a Basic Network Graph**

The code chunk below uses [`ggraph()`](https://ggraph.data-imaginist.com/reference/ggraph.html), [`geom-edge_link()`](https://ggraph.data-imaginist.com/reference/geom_edge_link.html) and [`geom_node_point()`](https://ggraph.data-imaginist.com/reference/geom_node_point.html) to plot a network graph.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

## 4.2 Changing the Default Network Graph Theme

We use `theme_graph()` to remove the x and y axes.

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

## 4.3 Changing the Colour of the Plot

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

## 4.4 Working with *ggraph*'s Layouts

**ggraph** supports many layout for standard used: star, circle, nicely (default), dh, gem, graphopt, grid, mds, spahere, randomly, fr, kk, drl and lgl.

The `layout` argument is used to define the layout to be used.

The code chunk below plots the network graph using Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

## 4.5 Modify Network Nodes

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

::: callout-note
-   *geom_node_point* is equivalent in functionality to *geo_point* of **ggplot2**.

-   It allows for simple plotting of nodes in different shapes, colours and sizes. In the code chunks above, colour and size are used.
:::

## 4.6 Modify Edges

In the code chunk below, the thickness of the edges is mapped with the `Weight` variable.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

::: callout-note
-   *geom_edge_link* draws edges in the simplest way - as straight lines between the start and end nodes.

-   In the example above, argument *`width`* is used to map the width of the line in proportional to the `Weight` attribute, and argument `alpha` is used to introduce opacity on the line.
:::

# 5. Creating Facet Graphs

There are three functions in **ggraph** to implement faceting, they are:

-   [`facet_nodes()`](https://r4va.netlify.app/chap27): edges are only draw in a panel if both terminal nodes are present here

-   [`facet_edges()`](https://ggraph.data-imaginist.com/reference/facet_edges.html): nodes are always drawn in al panels even if the node data contains an attribute named the same as the one used for the edge faceting

-   [`facet_graph()`](https://ggraph.data-imaginist.com/reference/facet_graph.html) faceting on two variables simultaneously

## 5.1 Working with *facet_edges()*

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

## 5.2 Working with *theme()*

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

## 5.3 A Framed Facet Graph

Adding frame to each graph.

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## 5.4 Working with *facet_nodes()*

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

# 6. Network Metric Analysis

## 6.1 Computing Centrality Indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

::: callout-note
-   `mutate()` of **dplyr** is used to perform the computation.

-   The algorithm used is the `centrality_betweenness()` of **tidygraph**.
:::

## 6.2 Visualising Network Metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

## 6.3 Visualising Community

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

# 7. Building Interactive Network Graph with *visNetwork*

-   [`visNetwork()`](http://datastorm-open.github.io/visNetwork/) is a R package for network visualization, using [vis.js](http://visjs.org/) javascript library.

-   `visNetwork()` function uses a nodes list and edges list to create an interactive graph.

    -   The nodes list must include an “id” column, and the edge list must have “from” and “to” columns.

    -   The function also plots the labels for the nodes, using the names of the actors from the “label” column in the node list.

To get starts, we prepare the data model.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

## 7.1 Plotting First Interactive Network Graph

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

## 7.2 Working with Layout

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

## 7.3 Working with Visual Attributes - Nodes

`visNetwork()` looks for a field called “group” in the nodes object and colour the nodes according to the values of the group field.

The code chunk below renames Department field to `group`.

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

visNetwork shades the nodes by assigning unique colour to each category in the `group` field.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## 7.4 Working with Visual Attributes - Edges

`visEdges()` is used to symbolise the edges:

-   The `arrows` argument is used to define where to place the arrow.

-   The `smooth` argument is used to plot the edges using a smooth curve.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## 7.5 Interactivity

`visOptions()` is used to incorporate interactivity features in the data visualisation.

-   The argument `highlightNearest` highlights nearest when clicking a node.

-   The argument `nodesIdSelection` adds an id node selection creating an HTML select element.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

[^1]

[^1]: This document was completed with reference to:

    -   Prof Kam's notes "27 Modelling, Visualising and Analysing Network Data with R": <https://r4va.netlify.app/chap27>
