---
title: "In class Ex04"
author: "Jia Peng Chua"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

We will launch the following packages in R.

```{r}
pacman::p_load(haven, SmartEDA, tidyverse, 
               tidymodels, ggridges, ggthemes,
               colorspace, ggdist)
```

::: callout-note
`tidyverse` is a collection of R packages for data science.

In `tidyverse`, `dplyr` is useful for data transformation. `ggplot2` is useful for visualisation.
:::

The code chunk below imports the *Exam_data.csv* file into R using the `read_csv()` function of *readr* package.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# 2. Box plot

```{r}
ggplot(data = exam_data,
       aes(y = CLASS,
           x = ENGLISH)) + 
    geom_boxplot() 
```

::: callout-note
Use `coord_flip()` to rotate the axis, or simply switch the variables in `x=` and `y=`.
:::

Boxplot is incomplete as it shows the summary statistics, instead of the distribution. To show distribution, ridgelines plot will help us visualise better.

```{r}
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges()
```

# 3. Additional Notes

```{r}
ggplot(exam_data, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

-   `geom_density_ridges()`: Extra geometric that supports the creation of ridge line

-   `scale_x_continuous()` and `scale_y_discrete()`: belong to *ggplot2*, for continuous and discrete values respectively

-   `expand = expansion(add = c(0.2, 2.6))`: makes the plot visually nicer

-   `fill = lighten("#7097BB", .3)`: fill each plot with a specific colour

-   `color = "white"`: border colour

-   Features of different packages can be combined

```{r}
ggplot(exam_data, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```

-   Dotplot provides details of the distribution

    -   also allows us to view the proportion of data, e.g. "Others" contains fewer datapoints

-   Ridgeline plot smoothens out the distribution
