---
title: "Take-Home Exercise 01 DRAFT 1"
author: "Jia Peng Chua"
date: "February 1, 2025"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Introduction

# 2. Getting Started

(intro and significance of data)

## 2.1 Loading the relevant libraries

The following R packages will be used:

-   `tidyverse`:

-   `performance`:

```{r}
pacman::p_load(tidyverse, performance, plotly,
               SmartEDA, easystats, gtsummary, ggstatsplot, ggridges, ggdist)
```

## 2.2 Data source and importing data

In this assignment, the "Heart Attack in Japan Youth Vs Adult" data will be used. The dataset is retrieved from [Kaggle](https://www.kaggle.com/datasets/ashaychoudhary/heart-attack-in-japan-youth-vs-adult?resource=download).

We import the dataset, `japan_heart_attack_dataset.csv`, using `read_csv()` of *readr* package into R.

```{r}
heart_attack <- read_csv("data/japan_heart_attack_dataset.csv")
```

# 3. Data Pre-Processing

View data

```{r}
summary(heart_attack)
```

Remove extra columns

```{r}
heart_attack <- heart_attack[, 1:17]
```

Check for duplicated values and missing values

```{r}
# Check for duplicated values
heart_attack[duplicated(heart_attack),]

# Check for missing values
any(is.na(heart_attack))
```

```{r}
heart_attack$Age_Group <- cut(heart_attack$Age, 
                              breaks = c(0, 25, 35, 45, 55, 65, 75, Inf), 
                              labels = c("Below 25", "25-34", "35-44", "45-54", "55-64", "65-74", "75+"), 
                              right = FALSE)
table(heart_attack$Age_Group)
```

```{r}
heart_attack$Heart_Attack_Occurrence <- ifelse(heart_attack$Heart_Attack_Occurrence == "Yes",1,0)

heart_attack
```

```{r}
heart_attack %>% 
    ExpData(type = 1)
```

```{r}
heart_attack %>% 
    ExpData(type = 2)
```

# 4. Exploratory Data Analysis

```{r}
model <- glm(Heart_Attack_Occurrence ~., family = binomial(link = 'logit'), data = heart_attack)
model
```

```{r}
cln_heart_attack <- check_collinearity(model)
plot(cln_heart_attack)
```

```{r}
check_norm <- check_normality(model)
plot(check_norm)
```

```{r}
heart_attack_1 <- heart_attack %>% filter(Heart_Attack_Occurrence == 1)
heart_attack_0 <- heart_attack %>% filter(Heart_Attack_Occurrence == 0)
```

```{r}
p1 <- ggplot(data = heart_attack, 
             aes(x= BMI, 
                 y= Cholesterol_Level)) +
  geom_point() +
  geom_smooth(method = lm, 
              linewidth = 0.5) +  

  ggtitle("Cholesterol Level vs Stress Levels")

p1
```

```{r}
bp <- heart_attack_1 %>%
  plot_ly(x = ~Stress_Levels, 
          y = ~BMI, 
          size = ~BMI, 
          color = ~Region,
          sizes = c(2, 100),
          frame = ~Age_Group, 
          text = ~Physical_Activity, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```

```{r}
ggplot(data=heart_attack, 
       aes(x = Stress_Levels)) +
  geom_histogram()
```

```{r}
ggplot(data=heart_attack, 
       aes(y = BMI,       
           x = Heart_Attack_Occurrence)) +    
  geom_boxplot()  
```

```{r}
ggplot(data = heart_attack, 
       aes(x = Cholesterol_Level, 
           y = Age)) +
  geom_point()  
```

```{r}
ggplot(data = heart_attack, 
       aes(y = Cholesterol_Level, x = Gender)) +
    geom_boxplot()
```

```{r}
ggplot(data = heart_attack, 
       aes(x = Age)) +
  geom_histogram(bins=20) +
    facet_wrap(~ Heart_Attack_Occurrence)
```

```{r}
ggplot(data = heart_attack, 
       aes(x = BMI, fill = Gender)) +
  geom_histogram(bins=20) +
    facet_wrap(~ Heart_Attack_Occurrence)
```

```{r}
ggplot(data = heart_attack_1, 
       aes(x = Heart_Rate, fill = Family_History)) +
  geom_histogram(bins=20) +
    facet_wrap(~ Age_Group)
```

```{r}
p1 <- ggplot(heart_attack_1,
       aes(x = Stress_Levels,
           y = Age_Group,
           fill = factor(stat(quantile)))) +
    stat_density_ridges(geom = "density_ridges_gradient",
                        calc_ecdf = TRUE,
                        quantile_lines = TRUE,
                        quantiles = 4) +
    scale_fill_viridis_d(alpha = 0.8) +
    xlab("Stress Level") + ylab("Age Group") +
    theme(plot.title = element_text(size = 10),
          legend.position = "none")+
    ggtitle(label = "Distribution of Stress Levels\nof Heart Attack Patients\nby Age Group")
```

```{r}
p2 <- ggplot(heart_attack_0,
       aes(x = Stress_Levels,
           y = Age_Group,
           fill = factor(stat(quantile)))) +
    stat_density_ridges(geom = "density_ridges_gradient",
                        calc_ecdf = TRUE,
                        quantile_lines = TRUE,
                        quantiles = 4) +
    scale_fill_viridis_d(alpha = 0.8) +
    xlab("Stress Level") +
    theme(plot.title = element_text(size = 10),
          legend.position = "none",
          axis.title.y = element_blank())+
    ggtitle(label = "Distribution of Stress Levels\nof Non-Heart Attack Patients\nby Age Group")
```

```{r}
p1 + p2
```
