---
title: "Take-Home Exercise 01"
author: "Jia Peng Chua"
date: "February 1, 2025"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Introduction

# 2. Getting Started

(intro and significance of data)

## 2.1 Loading the relevant libraries

The following R packages will be used:

-   `tidyverse`:

-   `performance`:

```{r}
pacman::p_load(tidyverse, performance, plotly)
```

## 2.2 Data source and importing data

In this assignment, the "Heart Attack in Japan Youth Vs Adult" data will be used. The dataset is retrieved from [Kaggle](https://www.kaggle.com/datasets/ashaychoudhary/heart-attack-in-japan-youth-vs-adult?resource=download).

We import the dataset, `japan_heart_attack_dataset.csv`, using `read_csv()` of *readr* package into R.

```{r}
heart_attack <- read_csv("data/japan_heart_attack_dataset.csv")
```

# 3. Data Pre-Processing

View data

```{r}
glimpse(heart_attack)
```

Remove extra columns

```{r}
heart_attack <- heart_attack[, 1:17]
```

Check for duplicated values and missing values

```{r}
# Check for duplicated values
heart_attack[duplicated(heart_attack),]

# Check for missing values
any(is.na(heart_attack))
```

# 4. Exploratory Data Analysis

```{r}
heart_attack$Heart_Attack_Occurrence <- ifelse(heart_attack$Heart_Attack_Occurrence == "Yes",1,0)

heart_attack
```

```{r}
model <- glm(Heart_Attack_Occurrence ~., family = binomial(link = 'logit'), data = heart_attack)
model
```

```{r}
cln_heart_attack <- check_collinearity(model)
plot(cln_heart_attack)
```

```{r}
check_norm <- check_normality(model)
plot(check_norm)
```

```{r}
p1 <- ggplot(data = heart_attack, 
             aes(x= BMI, 
                 y= Cholesterol_Level)) +
  geom_point() +
  geom_smooth(method = lm, 
              linewidth = 0.5) +  

  ggtitle("Cholesterol Level vs Stress Levels")

p1
```

```{r}
bp <- heart_attack %>%
  plot_ly(x = ~Stress_Levels, 
          y = ~Cholesterol_Level, 
          size = ~BMI, 
          color = ~Physical_Activity,
          sizes = c(2, 100),
          frame = ~Age, 
          text = ~Physical_Activity, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```

```{r}
ggplot(data=heart_attack, 
       aes(x = Stress_Levels)) +
  geom_histogram()
```

```{r}
ggplot(data=heart_attack, 
       aes(y = BMI,       
           x = Heart_Attack_Occurrence)) +    
  geom_boxplot()  
```

```{r}
ggplot(data = heart_attack, 
       aes(x = Cholesterol_Level, 
           y = Age)) +
  geom_point()  
```

```{r}
ggplot(data = heart_attack, 
       aes(y = Cholesterol_Level, x = Gender)) +
    geom_boxplot()
```

```{r}
ggplot(data = heart_attack, 
       aes(x = Age)) +
  geom_histogram(bins=20) +
    facet_wrap(~ Heart_Attack_Occurrence)
```

```{r}
ggplot(data = heart_attack, 
       aes(x = BMI, fill = Gender)) +
  geom_histogram(bins=20) +
    facet_wrap(~ Heart_Attack_Occurrence)
```

```{r}
ggplot(data = heart_attack, 
       aes(x = Age, y = Stress_Levels)) +
  geom_violin() +
    facet_wrap(~ BMI)
```
